%========== preamble ==========%

\newpage
\pagestyle{fancy}

\chapter*{Preamble}
In this document we will discuss the most important principles of algorithms
and datastructures in accordance with the syllabus of the course under the
same title at the University of Copenhagen, Department of Computer Science.

\section*{About the authors}
We're a small group of Computer Science students, attending the course
mentioned above at the time of writing this document as collaborative effort
to make a great disposition for the course exam, that will both increase our
own understanding of the subjects discussed as well as produce an invaluable
resource for others attending the course, or who wish to learn about
algorithms and datastructures in generel.

\section*{Why we made this document}
The document is was written in order to support our own understanding of the
course contents as well as aid us during the exam preparation time.

\section*{How to use this document}
The intention of the document is \textit{not} to be a fullfledged introduction
to algorithms and datastructure that you can just pick up and start learning
from. As such it requires at least an accompanying book, or that you are
attending lectures on the subject, or some other means that provide a base
knowledge.

The document is merely a tool for looking up things quickly, and defines the
subjects in a concise manner, hence no in-depth discussion of any subject is
to be found in this document. The sole purpose of the document is precisely
that; a resource for quickly looking up the concepts discussed presented at
lectures, or in a book.

\section*{Recommendations}
As far as we know the book \textit{Introduction to Algorithms} by Thomas H.
Cormen, Charles E. Leiserson, Ronald L. Rivest and Clifford Stein, is pretty
much the go-to choice if you want to study algorithms, and since this was our
textbook during the course this is what we'll recommend for you.

