%========== Priority Queues ==========%

\chapter{Priority Queues}
\label{ch:priorityqueues}

\textbf{Relevant Assignment} Problem ?-?\\\\
\textbf{Pensum}CLRS Ch. 6\\\\
\textbf{Keywords} Min- and maxheap, stacks
\vspace{1in}

\noindent A priority queue is an abstract data structure, which sorts and
maintains a set of data in such a way, that the data is prioritized. It is an
important distinction to make that a priority queue isn't a particular data
structure, but rather a class of data structures. Data structures that work
well with this notion include stacks, heaps, self-organizing lists, etc.
We will focus mainly on heaps.

\section{Heap}
A heap is a nearly complete $n$-nary tree, we will concern ourselves only with
binary heaps - and as such, a binary heap is a nearly complete binary tree. A
heap, in the sense of priority queues implements a set $S$ of elements, in
which each element has an associated key. The actual structure of a heap is
simply an array, and so a heap is defined by the operations performed on an
array, rather than the data structure itself. The reason a heap is a
tree-structure is a visualization of the data, but not reflected by the data
structure in any way.

Visualizing the array of a heap as a tree, we have that the root of the tree
corresponds to the first element of the array $i = 0$ (zero-indexed). The
parent of any node $n$ is given by $\lfloor n/2 \rfloor$. The children are
given by $2n$ and $2n + 1$, for left and right, respectively.
\\\\
A heap, however, is not a priority queue in and of itself, it must adhere to
a set of properties, and that is either the max- or min-heap properties.
\begin{description}
	\item \textbf{Max-Heap} The key of a node is greater than or equal to the
keys of its children.
	\item \textbf{Min-Heap} The key of a node is less than or equal to the
keys of its children.
\end{description}
As the max- and min heaps are defined by the procedures performed on them, we
must show that the their property is maintained by each of these procedures.

% TODO: prove that all procedures maintain the max- or min-heap property

\section{Stack}
% extra stuff, don't focus on this, I think.
A stack is one of the simplest queuing data structures. [...]

